Regression Cube Analysis of Cohort Study Data

### 2015-03-26 draft.tex
added details on the application section

Both data sets are unusual for epidemiological analysis regarding their feature extend.
%%
Usually, only a few features depicting a hypotheses are compiled into a data set to assess them using statistical tools.
%%
Our method focuses on data exploration and knowledge extraction and requires a wide scope of sociodemographic, medical and lifestyle features.
%%
\subsection{Participants, Setup and Procedure}
%%
To assess the ability of a system to discover knowledge is hard to measure.
%%
Lam et al. \cite{Lam2012} propose for this purpose the \emph{Visual Data Analysis and Reasoning} (\emph{VDAR}) technique, which is focused on the characterization of a systems' ability to generate hypotheses and explore the data in order to extract information.
%%
\emph{VDAR} can be carried out using case studies using thinking-aloud techniques to comprehend the reasoning and thought process of the user.
%%
We conducted a study with two participants, who also co-authored this publication.
%%
UN is a data scientist who analyzed the hepatic steatosis data set by extracting decision rules.
%%
KH is a radiologist (10 years of experience) with specialization in epidemiological research and is also responsible for the SHIP MRI acquisition.
%%
\paragraph{Setup and Procedure.}
%%
One major advantage of focusing on web-based research was the possibility to bridge the geographical gap towards the epidemiological domain experts.
%%
We conducted a web-based research by using an online-meeting software, which features voice chat as well as screen sharing.
%%
Starting an analysis using these techniques take about 5-10 minutes of setup time.

The sessions started with an initial overview of the system, showcasing its features and functionality.
%%
After that the experts use the system on their own computers.
%%
The screen-sharing function was still used to observe the actions of the expert.
%%
All sessions were video recorded to be processed later on.
%%
\subsection{The Breast Fat Data Set}
The breast fat data set was compiled to find associations between parenchyma tissue proportion in the female breast towards other features in the data.
%%
The ratio between parenchyma and cellular connective tissue (breast density) has been shown to be associated with breast cancer.
%%
Studies describe a four to five times increased risk of getting breast cancer for participants with a breast density above 50\% \cite{Mccormack2006}.

The data comprises 1.186 subjects (368 from \texttt{SHIP-2}, 818 from \texttt{SHIP-TREND-0} cohort).
%%
It contains 230 dimensions, holding information about
\begin{itemize}
	\item \emph{somatometric features}, e.g., body size, weight or BMI, 
	\item \emph{lifestyle features}, e.g., alcohol/tobacco consume, 
	\item \emph{personal history}, e.g., occupation, marital status,
	\item \emph{medical history}, e.g., current or prior diseases or laboratory values, such as cholesterol levels,
	\item \emph{women specific features}, e.g., number of born children, contraception type or hormone replacement therapy and
	\item \emph{mammography features} derived from the image data.
\end{itemize}
%%
The latter were derived using a level set based segmentation \cite{Ivanovska2014} of MRI image data for each subject.
%%
\com{Check this with Mrs. Hegenscheid, probably most of them were segmented by hand!}
%%
%\com{Describe Parenchym tissue and its relation to breast cancer.}
%\com{Describe how the data was acquired and preprocessed}

@@

In \cite{Niemann2014}, the authors show that women and men exhibit differences in the class distribution of the liver fat concentration.
%%
Also, for women they identify that there is a association between age and liver fat concentration.
%%
They compute an appropriate cut-off value of 52 years for which the class distribution is most homogeneous within the resulting subsets.
%%
Based on this observations, we perform our analysis on three sub-populations, $PartitionF$, $PartitionM$ and \emph{$F:age\>52$}.
%%
\subsection{Case 1: Hypothesis-free Analysis of the Breast Cancer Data Set}
%%
\subsection{Case 2: Hypothesis-driven Analysis of the Hepatic Steatosis Data Set}
%%
\subsection{Further Feedback and Lessons Learned} \label{Lessons Learned}
%%
\com{Feedback from the evaluation goes here.}
\com{Time-aspect critical, interactive analysis requires for fast response. The method needs to be speeded up.}

\section{Summary and Conclusion}
We presented a technique for knowledge discovery in cohort study data sets with user-defined target features.
%%
Dimension reduction using the target restricts the analysis to most important features.
%%
\emph{Hypothesis-free} analysis employs default regression models.
%%
Model expert knowledge using regression formulas allow for \emph{hypothesis-based} investigation.
%%
A 3D \emph{Regression Cube} allows to assess hotspots in the analysis by abstracting regression models using a quality-of-fit measure.
%%
These can then be analyzed further using the 2D mosaic plot for each cube slice.
%%
Details on demand for each model allow for detailed assessment of regression models.
%%
We successfully applied the approach to find correlations in a hepatic steatosis as well as a breast cancer data set.
%%
The method was well received by our clinical partners, triggering detailed investigations of the findings.

>>>>

Both data sets are unusual for epidemiological analysis regarding their feature extend.
%%
Usually, only a few features depicting a hypotheses are compiled into a data set to assess them using statistical tools.
%%
Our method focuses on data exploration and knowledge extraction and requires a wide scope of sociodemographic, medical and lifestyle features.
%%
\subsection{Participants, Setup and Procedure}
%%
To assess the ability of a system to discover knowledge is hard to measure.
%%
Lam et al. \cite{Lam2012} propose for this purpose the \emph{Visual Data Analysis and Reasoning} (\emph{VDAR}) technique, which is focused on the characterization of a systems' ability to generate hypotheses and explore the data in order to extract information.
%%
\emph{VDAR} can be carried out using case studies using thinking-aloud techniques to comprehend the reasoning and thought process of the user.
%%
We conducted a study with two participants, who also co-authored this publication.
%%
UN is a data scientist who analyzed the hepatic steatosis data set by extracting decision rules.
%%
KH is a radiologist (10 years of experience) with specialization in epidemiological research and is also responsible for the SHIP MRI acquisition.
%%
\paragraph{Participants, Setup and Procedure.}
%%
One major advantage of focusing on web-based research was the possibility to bridge the geographical gap towards the epidemiological domain experts.
%%
We conducted a web-based research by using an online-meeting software, which features voice chat as well as screen sharing.
%%
Starting an analysis using these techniques take about 5-10 minutes of setup time.

The sessions started with an initial overview of the system, showcasing its features and functionality.
%%
After that the experts use the system on their own computers.
%%
The screen-sharing function was still used to observe the actions of the expert.
%%
All sessions were video recorded to be processed later on.
%%

Domain expert for the breast fat data set analysis is KH, a clinician (10 years of experience) with focus on epidemiological research.
%%
KH is a radiologist, responsible for the SHIP-MRI acquisition and also responsible of the mammography analysis.

The hepatic steatosis data set is analyzed by UN, a data scientists responsible for prior analysis of the data with.
%%

TI, a statistician with focus on epidemiology (\com{TODO} years of experience) assesses the statistical reliability of the tool and the underlying methods without a focus on a specific data set.
%%
\subsection{The Breast Fat Data Set}
The breast fat data set was compiled to find associations between parenchyma tissue proportion in the female breast towards other features in the data.
%%
The ratio between parenchyma and cellular connective tissue (breast density) has been shown to be associated with breast cancer.
%%
Studies describe a four to five times increased risk of getting breast cancer for participants with a breast density above 50\% \cite{Mccormack2006}.

The data comprises 1.186 subjects (368 from \texttt{SHIP-2}, 818 from \texttt{SHIP-TREND-0} cohort).
%%
It contains 231 dimensions, holding information about
\begin{itemize}
	\item \emph{somatometric features}, e.g., body size, weight or BMI, 
	\item \emph{lifestyle features}, e.g., alcohol/tobacco consume, 
	\item \emph{personal history}, e.g., occupation, marital status,
	\item \emph{medical history}, e.g., current or prior diseases or laboratory values, such as cholesterol levels,
	\item \emph{women specific features}, e.g., number of born children, contraception type or hormone replacement therapy and
	\item \emph{mammography features} derived from the image data.
\end{itemize}
%%
The latter were derived using a level set based segmentation \cite{Ivanovska2014} of MRI image data for each subject.
%%
\com{Check this with Mrs. Hegenscheid, probably most of them were segmented by hand!}
%%
%\com{Describe Parenchym tissue and its relation to breast cancer.}
%\com{Describe how the data was acquired and preprocessed}

@@

In \cite{Niemann2014}, the authors show that women and men exhibit differences in the class distribution of the liver fat concentration.
%%
Also, for women they identify that there is a association between age and liver fat concentration.
%%
They compute an appropriate cut-off value of 52 years for which the class distribution is most homogeneous within the resulting subsets.
%%
Based on this observations, we perform our analysis on three sub-populations, $PartitionF$, $PartitionM$ and \emph{$F:age\>52$}.
%%
\subsection{Case 1: Hypothesis-free Analysis of the Breast Cancer Data Set}
%%
The analysis goal is to find relationships on the breast cancer data using features derived from the mammography analysis.
%%
Relationships between the share of parenchyma tissue on the overall breast volume are of high interest \cite{Mccormack2006}.
%%

\subsection{Case 2: Hypothesis-driven Analysis of the Hepatic Steatosis Data Set}
%%
\subsection{Further Feedback and Lessons Learned} \label{Lessons Learned}
%%
\textbf{Extracted Hypotheses have to investigated further.}
\com{Tills Feedback - Regression analysis also incorporates assumptions about the data, }
%%

\textbf{Overview Visualizations gain importance with increasing feature count.}
%%
While the classical epidemiological workflow (see Sec.~\ref{sec:Background}) represents the foundation of the research in this area, explorative analysis based on the data gains importance.
%%
Incorporating data mining algorithms with statistically established and overview visualizations allow for reasoning about risk factors using underlying epidemiological data sets.
%%
One major advantage we observed is the confidence of an expert in the results he or she actually observe.
%%
Their participation in the analysis using human pattern-detection and expert knowledge is for experts preferred over results derived using a black-box method.
%%
By observing expected correlations matching the expert knowledge strengthens the confidence in the method and subsequently in the hypothesis generated from unexpected relationships.
%%
%\com{Include the human into the analysis provides more confidence into the result than data mining algorithms. People believe more in what they see, rather than in what they are presented by an algorithm. Regression cubes incorporate this by showing known associations as well as new ones. Preattentive feature display (either through fast slicing through a cube or by showing hot spots) helps.}

%%
\textbf{Using non-discretized features reduces information bias.}
%%
\com{Comment on the lesson learned from the last VAST Paper - "Usage of different categorizations depending on expected outcome." Now we have continuos data.}
%%

\textbf{Support for different subject groups.}
\com{Include this?}
%%
\com{Feedback from the evaluation goes here.}
\com{Time-aspect critical, interactive analysis requires for fast response. The method needs to be speeded up.}

\section{Summary and Conclusion}
%%
We presented a technique for knowledge discovery in cohort study data sets with user-defined target features.
%%
Dimension reduction using the target restricts the analysis to most important features.
%%
\emph{Hypothesis-free} analysis employs default regression models.
%%
Model expert knowledge using regression formulas allow for \emph{hypothesis-based} investigation.
%%
A 3D \emph{Regression Cube} allows to assess hotspots in the analysis by abstracting regression models using a quality-of-fit measure.
%%
These can then be analyzed further using the 2D mosaic plot for each cube slice.
%%
Details on demand for each model allow for detailed assessment of regression models.
%%
We successfully applied the approach to find correlations in a hepatic steatosis as well as a breast cancer data set.
%%
The method was well received by our clinical partners, triggering detailed investigations of the findings.
